---
title: "Preliminary examination"
author: "Amey Joshi"
date: "29/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(MASS)
library(lattice)

training.data <- read.csv("train.csv")
```

```{r}
data_availability <- function(column.name, show.grp.size = TRUE) {
    num.na <- nrow(training.data[is.na(training.data[[column.name]]), ])
    cat(
        paste(
            "The proportion of values for which",
            column.name,
            "is NA is",
            round(num.na / nrow(training.data), 4),
            "\n"
        )
    )
    
    if (show.grp.size) {
        grp.size <- tapply(training.data$SalePrice,
                           training.data[[column.name]],
                           length)
        cat(paste("Group size of", column.name, "are:\n"))
        print(grp.size)
        cat("\n")
    }
    
    rm(num.na)
}

check_group_means <- function(column.name, run.anova = TRUE) {
    grp.means <- tapply(training.data$SalePrice,
                        training.data[[column.name]],
                        mean)
    cat(paste("Group means of", column.name, "are:\n"))
    print(grp.means)
    
    if (run.anova) {
        anova.res <-
            oneway.test(training.data$SalePrice ~ training.data[[column.name]])
        print(anova.res)
        rm(anova.res)
    }
    
    boxplot(training.data$SalePrice ~ training.data[[column.name]],
            xlab = column.name,
            ylab = "Sale Price")
}
```

## Observations
The observations listed in this section are substantiated in the rest of the 
analysis.

1.  Sale price for different sales condition is different. The categories 
    'Abnorml' and 'Family' are statistically similar.
2.  It might be useful to collapse sale type into 'Normal', 'New' and 'Others'.
    Sale type and sale condition seem to be closely related to each other. We can
    think of retaining only one of them.
3.  We need both Zoning classification and the dwelling type.
4.  Neither lot frontage nor lot area are strongly correlated with sale price.
5.  Street and Alley may not be a good predictor of sale price.
6.  Lot shape may be a predictor of sale price.
7.  Land contour too may be a predictor of sale price.
8.  Utilities cannot be used used to predict sale price.
9.  Lot configuration may be a good predictor but it may be helpful to collapse
    a few levels.
10. Land slope is not a candidate predictor of sale price.  
11. Neighborhood is a predictor of sale price.
12. Proximity to various conditions (Condition1) may be good predictors of sale 
    price. However, Condition2 might not be very helpful. The $9$ levels of 
    Condition1 can be coalesced into three.
13. BldgType may be a predictor of sale price. Its $5$ levels can be coalesced
    into $2$.
14. HouseStyle may be a predictor of sale price.
15. Overall quality and condition are strong predictors of sale price.
16. YearBuilt and YearRemodAdd are useful to predict sale price. However, we
    must transform them to intervals of age and 'years since last remodeling'
    before we can use them.
17. Roof style may be used as a predictor. Roof material can be used only if
    we collapse its eight levels into two.
18. The variables Exterior1st and Exterior2nd are hard to use without merging
    pairs of levels.
19. Masonry veneer type and manonry veneer area may be candidate predictors of
    sale price. However, they come into play if veneer type is not 'None'. If at
    all we choose to select this variable in our model we may need separate model
    when veneer type is 'None' and when it is not. The correlation coefficients 
    of veneer area and sale price are not too dissimilar in the cases when veneer
    type is not 'None'. Perhaps, the type of veneer is may not strongly influence
    sale price.
20. External quality and condition are very strong indicators of sale price.  
21. Basement related variables:
    + Basement quality and condition strongly influence sale price.
    + Basement exposure strongly influences sale price.
    + Basement finishing is a candidate influencer. However, a few levels can 
      be grouped together.
    + Of the four areas related to the basement only the total area is strongly
      correlated with the sale price.
22. Heating is not a strong indicator of sale price but HeatingQC is.     
23. Presence or absence of central air-conditioning strongly influences sale 
    price.
24. Electrical system is also a strong candidate to determine sale price.
25. Of the four area variables X1stFlrSF, X2ndFlrSF, LowQualFinSF and GrLivArea, 
    the last one is most strongly correlated with sale price.
26. The number of full bathrooms is the best indicator of sale price among, the
    four variables related to bathrooms.
27. The sale price is sharply different for different number of bedrooms above
     ground. However, the relation between the two is counter-intuitive. The
     mean sale price for houses with no bedrooms above ground is among the 
     highest. The group also has a small sample and a very high variance. I 
     think, to the extent possible, I should avoid this variable in a model.
28. The quality of the kitchens matter more than their number. 
29. The number of rooms above ground does seem to predict the sale price. It does
    not have the anomalies that the number of bedrooms had.
30. The variable 'Functional' is a predictor of sale price.
31. The quality of the fireplaces is a good predictor of sale price than just 
    their number.
32. Garage related variables:
    + Garage type does resolve sale price but two of six types have statistically
      similar mean sale price.
    + The age of the garage may have only a weak influence on the sale price.
      Additionally, the are a few NAs for this attribute.
    + Garage finish is a good predictor of sale price.
    + The capacity of the garage resolves the sale price. However, the variation
      is not monotonic. One house with a capacity for four cars was sold at a 
      price much lower compared to others. The data is also not eenly distributed
      across its levels.
    + The quality of the garage may predict sale price. However, two of its levels
      have similar group means.
    I think that the condition of the garage cannot be different from the rest
    of the house. Therefore, we need not consider them unless we have to.
33. Paved drive is a predictor of sale price.
34. More than half the houses do not have a wood deck. Further, for the ones
    that have it, the correlation between its area and the sale price is not
    very significant. We may, therefore, drop it from the first step of our 
    analysis.
35. Variable related to porch:
    + More than a quarter of the houses do not have an open porch. For the ones
      that have it, the correlation between its area and the sale price is not 
      very significant. We may not consider this variable in the first iteration
      of our model.
    + A very large number of houses do not have an enclosed porch. Therefore, we
      do not consider EnclosedPorc.
    + Very few houses have a three seasons porch or a screened porch. Therefore,
      we do not consider these variables in the first iteration of the model.
    Only high-end houses have a porch. It may be a differentiator only for them.
    Unless we build a separate model for high-end houses, we may not use these 
    variables.
36. Pool related variables:    
    + A very large number of houses does not have a pool.
    + Pool quality data is available only for $7$ houses.
37. The four levels of the Fence variable do not resolve the sale price data 
    well enough.
38. Only $3.7\%$ cases have miscellaneous features. Therefore, the variables
    'MiscFeature' and 'MiscVal' may not be of use in our model.
39. The month and year a house was sold may have an effect on the sale price.
    However, the effect may be more because of economic conditions prevalent in
    the month than anything related to the house.
    
All these observations are made with the simple assumption that an attribute is 
a candidate predictor if price changes as the attribute value changes. We may
be wrong in assuming it. For a correlation between an attribute and sale price
may be spurious.
    
## Condition of sale
We first find the distribution of 'condition of sale'
```{r}
with(training.data,
     tapply(SalePrice, SaleCondition, function(x) {
       round(length(x)/nrow(training.data), 4)
     }))
```

A significant percentage of sales were normal. Is there a statistically
significant difference in their sale price? We first view the data
```{r}
with(training.data, boxplot(SalePrice ~ SaleCondition))
```

The data shows a large number of outliers even in the 'normal' case. Further,
all outliers are at the higher end of the sale price. Visually, it is hard to
tell if the sale price depends on the sale condition. We run an ANOVA to test it.

```{r}
oneway.test(SalePrice ~ SaleCondition, training.data)
```

The null hypothesis of the one way test is that all the means are equal. The 
test concludes that they are not all equal. Are some of them equal? We first get
the means.
```{r}
with(training.data,
     tapply(SalePrice, SaleCondition, mean))
```

Of these, the 'Abnorml' [sic] and 'Family' means are close by. Let us check if
they are statistically similar.
```{r}
oneway.test(SalePrice ~ SaleCondition, 
            data = training.data[training.data$SaleCondition %in% c("Abnorml", "Family"), ])
```

A $p$-value of $82\%$ suggests that we cannot reject the null hypothesis. The
data suggests that the means are similar. It is possible that 'Family' trades
happen during an abnormal situation and vice versa.

## Sale type
The distribution of sale type is
```{r}
with(training.data,
     tapply(SalePrice, SaleType, function(x) {
       round(length(x)/nrow(training.data), 4)
     }))
```

Is the sale type correlated with the sale condition?
```{r}
with(training.data, 
     table(SaleType, SaleCondition))
```

A large number of WD (Warranty Deed - Conventional) sale type are indeed normal.
The words normal and conventional are synonymous.


The mean sale price across the types is
```{r}
with(training.data,
     tapply(SalePrice, SaleType, mean))
```

It is tempting to conclude that the means are different until one also realizes
that there are only two transactions with sale type 'Cond'.

Does sale type determine sale price? 
```{r}
with(training.data,
     boxplot(SalePrice ~ SaleType))
```

The 'WD' class shows a large number of outliers. But this class also tend to 
have a 'normal' sale condition.

Let us check if the mean sale price is statistically distinct across sale types.
```{r}
oneway.test(SalePrice ~ SaleType, training.data)
```

What was suspected from the box-plot is substantiated by ANOVA. Instead of
looking at all the sale types, what if we club all non-WD together. We create
a new dataset with this change.
```{r}
temp.ds <- training.data[, c("SaleType", "SalePrice")]
temp.ds$SaleType.1 <- ifelse(training.data$SaleType == "WD", "WD", "Non-WD")
oneway.test(SalePrice ~ SaleType.1, data = temp.ds)
```

The 'WD' and 'non-WD' classes are indeed different. Their class means are
```{r}
with(temp.ds,
     tapply(SalePrice, SaleType.1, mean))
```

The distribution of data is
```{r}
with(temp.ds,
     boxplot(SalePrice ~ SaleType.1))
```

Among the 'Non-WD' is 'New' sufficiently different?
```{r}
temp.ds$SaleType.2 <- ifelse(training.data$SaleType == "WD", 
                             "WD",
                             ifelse(training.data$SaleType == "New", 
                                    "New", 
                                    "Others"))
oneway.test(SalePrice ~ SaleType.2, data = temp.ds)
```

It makes sense the separate "New" from the "Non-WD". The new class means are
```{r}
with(temp.ds,
     tapply(SalePrice, SaleType.2, mean))
```

The distribution of data is
```{r}
with(temp.ds,
     boxplot(SalePrice ~ SaleType.2))
```

How different are "Others" from "WD"?
```{r}
oneway.test(SalePrice ~ SaleType.2, data = temp.ds[temp.ds$SaleType.2 %in% c("Others", "WD"), ])
```

They differ but not as significantly as from "New".
```{r}
rm(temp.ds)
```

## Type of dwelling
A distribution of sale price across dwelling types is
```{r}
with(training.data,
     boxplot(SalePrice ~ MSSubClass))
```

Sale price seems to be strongly dependent on the type of the dwelling. The 
differences are visually conspicuous enough to not require a confirmation 
through ANOVA.

We will also look at the number of sales in each category.
```{r}
with(training.data,
     tapply(SalePrice, MSSubClass, function(x) {
       round(length(x)/nrow(training.data), 4)
     }))
```

Unlike the case of sale condition and sale type, dwelling type is not dominated
by one class alone.

## Zoning classification
```{r}
with(training.data,
     boxplot(SalePrice ~ MSZoning))
```


We will also look at the number of sales in each category.
```{r}
with(training.data,
     tapply(SalePrice, MSZoning, function(x) {
       round(length(x)/nrow(training.data), 4)
     }))
```

Most of the sales have happened in RL (Residential Low density) and RM (Residential
Medium density) zones.

How are the dwelling types distributed across zones?
```{r}
tbl.1 <- with(training.data,
              table(MSSubClass, MSZoning))
```

Are the classes 'RL' and 'RM' different? To check that we will first find the
distribution of the number of sales in 'RL' across MSSubClass. We will then run 
a 'goodness of fit' test to check if the 'expected distribution' that mimics that 
of 'RL' is similar to the actual 'RM' data.
```{r}
fisher.test(tbl.1[, c(4, 5)], simulate.p.value = TRUE)
```

The $p$-value indicates that the two distributions are not related. Let us also
check their correlation.
```{r}
cat(paste("Correlation between RL and RH is", 
          round(cor(tbl.1[, 4], tbl.1[, 5]), 4), "\n"))
rm(tbl.1)
```

Thus, we need both Zoning classification and the dwelling type.

## Lot frontage
It is the linear feet of road connecting the property. Let us summarize the
lot frontage data.
```{r}
summary(training.data$LotFrontage)
boxplot(training.data$LotFrontage,
        main = "Lot frontage")
```

Is there a correlation between lot frontage and sale price? Before we find it
we must remove the NAs.
```{r}
cc.lf <- complete.cases(training.data[, c("LotFrontage", "SalePrice")])
X <- training.data[cc.lf, c("LotFrontage", "SalePrice")]
cat(paste("Correlation between lot frontage and sale price is", 
          round(with(X, cor(LotFrontage, SalePrice)), 4),
          "\n"))
```

There is some correlation between the two. Let us visualize the data.
```{r}
with(X, plot(LotFrontage, 
             SalePrice, 
             xlab = "Lot frontage",
             ylab = "Sale price",
             main = "Sale price v Lot frontage", 
             cex = 0.1))
rm(cc.lf, X)
```

It is difficult to imagine a linear model predicting sale price using lot
frontage. Before leaving the variable. let us examine the histogram of the lot
frontage.
```{r}
truehist(training.data$LotFrontage, xlab = "Lot frontage")
```

## Lot area
Lot area is the size of the land in square feet. Its summary is
```{r}
summary(training.data$LotArea)
truehist(training.data$LotArea, xlab = "Lot area")
```

The lot area data is highly skewed. There are a very large number of lots with
a size around the median and mean (~ $10,000$ square feet) and a few with very
large size. Let us examine the correlation between lot area and sale price.
```{r}
cat(paste("Correlation between lot area and sale price is", 
          round(with(training.data, cor(LotArea, SalePrice)), 4),
          "\n"))
```

It is interesting to note that two variables do not have a very high correlation.
Perhaps, the land size is a poor predictor of the house price. A plot of sale 
price against the lot area is
```{r}
with(training.data,
     plot(LotArea,
          SalePrice,
          xlab = "Lot area",
          ylab = "Sale price", 
          cex = 0.1
          ))
```

## Street
There are only two levels to the variable 'Street', namely 'Grvl' for graveled
and 'Pave' for paved. Let us examine the mean sale price for these factors.
```{r}
with(training.data,
     tapply(SalePrice, Street, mean))
```

On of them is almost $50\%$ higher than the other. Let us confirm that they
are significant using an ANOVA.
```{r}
oneway.test(SalePrice ~ Street, data = training.data)
```

Interestingly, ANOVA does not tell that the two means are statistically 
different. This perhaps indicates that Street might not be a good predictor of
sale price.

We repeat the analysis for the variable 'Alley' which has the same two levels.
```{r}
with(training.data,
     tapply(SalePrice, Alley, mean))
```

The difference is not as sharp as it is in the case of Street. However, if we 
run an ANOVA
```{r}
oneway.test(SalePrice ~ Alley, data = training.data)
```

We realize that Alley matters. The two group means are statistically different.

It is a little baffling that Street and Alley should be so different. Let us 
look a little deeper in the data. We first check the number of NAs.
```{r}
data_availability("Street")
data_availability("Alley")
```

A very large proportion of cases do not have an Alley. Therefore, the statistical
difference in the group means of the levels of Alley may not be useful to predict
the sale price.

## Lot shape
Does regularity of lot shape matter. The group means are
```{r}
with(training.data, 
     tapply(SalePrice,
            LotShape,
            mean))
```

The are distinct. Yet we would like to see the box plot of sale price for each
type of lot shape.
```{r}
with(training.data,
     boxplot(SalePrice ~ LotShape))
```

Are the differences in the means statistically significant?
```{r}
oneway.test(SalePrice ~ LotShape, data = training.data)
```

ANOVA indicates that they are different.

## Land contour
There are four levels of land contour. Their group means are
```{r}
with(training.data, 
     tapply(SalePrice,
            LandContour,
            mean))
```

Their box plot is
```{r}
with(training.data,
     boxplot(SalePrice ~ LandContour))
```

Let us run an ANOVA to compare the group means.
```{r}
oneway.test(SalePrice ~ LandContour, data = training.data)
```

The difference are indeed significant.

## Utilities
Do utilities matter? Do we have enough data?
```{r}
data_availability("Utilities")
```

With just one case available for 'NoSeWa' and all others for 'AllPub' we cannot
use this variable to predict sale price.

## Lot configuration
We first check that we have enough data.
```{r}
data_availability("LotConfig")
```

Although the attribute has data there are only a few cases with FR2 and FR3. We
may want to combine them at a later stage. The group means, ANOVA and box plots 
are
```{r}
check_group_means("LotConfig")
```

The differences in the means are statistically siginificant. Yet, the mean of
FR3 and FR2 are far apart. In fact, 'Corner', 'FR2' and 'Inside' seem be to closer
to each other. Even the box plot seems to suggest the same. We should, therefore,
examine the following possibilities:

1. Combine 'Corner', 'FR2' and 'Inside' in a single category 'CFI'.
2. Keep 'Inside' separate in view of the large number of outliers.

## Land slope
There are three levels of land slope. We first check if there is enough data
for this attribute.
```{r}
data_availability("LandSlope")
```

The group means, ANOVA and box plots are
```{r}
check_group_means("LandSlope")
```

The differences between the group means are not statistically significant. 

## Neighborhood
There are $25$ levels to this attribute. We check the characteristics of the
data.
```{r}
data_availability("Neighborhood")
```

The group means, ANOVA and box plots are
```{r}
check_group_means("Neighborhood")
```

Not surprisingly, neighborhood is a predictor of sale price. 

## Proximity to various conditions
We will study the two conditions together.

```{r}
data_availability("Condition1")
data_availability("Condition2")
```

The data is quite skewed toward normal conditions. Further, almost all data for
'Condition2' is normal making it ineffective for prediction. Let us check if
deviation from normal changes the sale price. We will first analyze with all 
levels of 'Condition1'. The levels of the 'Condition1' (and 'Condition2')
attribute are

- Artery	Adjacent to arterial street
- Feedr	    Adjacent to feeder street	
- Norm	    Normal	       
- RRNn	    Within 200' of North-South Railroad 
- RRAn	    Adjacent to North-South Railroad 
- PosN	    Near positive off-site feature--park, greenbelt, etc. 
- PosA	    Adjacent to postive off-site feature 
- RRNe	    Within 200' of East-West Railroad 
- RRAe	    Adjacent to East-West Railroad

```{r}
check_group_means("Condition1")
```

The conditions which tend to reduce the price are
- Artery
- Feedr
- RRAe

```{r}
oneway.test(SalePrice ~ Condition1, 
            data = training.data, 
            subset = Condition1 %in% c("Artery", "Feedr", "RRAe"))
```

The null hypothesis, that the differences between group means are not 
statistically different, cannot be rejected.

The conditions which probably have no effect are
- Norm (by default)
- RRAn
- RRNe

```{r}
oneway.test(SalePrice ~ Condition1, 
            data = training.data, 
            subset = Condition1 %in% c("Norm", "RRAn", "RRNe"))
```

These three conditions can also be grouped together.

The conditions which tend to increase the price are
- PosA
- PosN
- RRNn

```{r}
oneway.test(SalePrice ~ Condition1, 
            data = training.data, 
            subset = Condition1 %in% c("PosA", "PosN", "RRNn"))
```

These three levels too can be coalesced.

## Building type
There are $5$ levels of this attribute. The characteristic of the data are
```{r}
data_availability("BldgType")
```

The group means, ANOVA and box plots are
```{r}
check_group_means("BldgType")
```

We may have scope to group the attributes. For example, the group means of 
1Fam and TwnHse are quite close. The group means of the other three attributes
too are close enough. It is tempting to check this possiblity right away.
```{r}
oneway.test(SalePrice ~ BldgType, 
            data = training.data, 
            subset = BldgType %in% c("1Fam", "TwnhsE"))
```

The group means can be considered to be equal.

```{r}
oneway.test(SalePrice ~ BldgType, 
            data = training.data, 
            subset = BldgType %in% c("2fmCon", "Duplex", "Twnhs"))
```

One again, the group means can be considered to be equal.

## House style
There are $8$ levelsof house style. The characteristics of the data are
```{r}
data_availability("HouseStyle")
```

The group means, ANOVA and box plots are
```{r}
check_group_means("HouseStyle")
```

## Overall quality and condition
These are numeric attributes. We first check their availablity.
```{r}
data_availability("OverallQual")
data_availability("OverallCond")
```

Is there a correlation between these attributes?
```{r}
with(training.data,
     plot(OverallQual, 
          OverallCond, 
          xlab = "Overall quality", 
          ylab = "Overall condition", 
          cex = 0.1))
```

Visually, there is none. Yet. we support our observation by calculating the
correlation coefficient.
```{r}
cat(paste("Correlation between overall quality and overall condition is",
          round(with(training.data, cor(OverallQual, OverallCond)), 4),
          "\n"))
```

Indeed, the two variables are poorly correlated. Let us now check if they are
good predictors of sale price.
```{r}
check_group_means("OverallQual")
```

The overall quality, not surprisingly, is a very strong indicator of sale
price.

```{r}
with(training.data,
     tapply(SalePrice, OverallCond, mean))
with(training.data,
     boxplot(SalePrice ~ OverallCond))
```

We are unable to carry out a full ANOVA because some levels lack data. Let us
check if we can get results by dropping them.
```{r}
oneway.test(SalePrice ~ OverallQual, 
            data = training.data,
            subset = !(OverallQual %in% c(1, 2)))
```

The group means are indeed different.

## Original construction date and remodel date
These are not dates but years. Further, the remodel date is the same as original
construction date if the house has not been remodeled. 
We first check their availablity.
```{r}
data_availability("YearBuilt", show.grp.size = FALSE)
data_availability("YearRemodAdd", show.grp.size = FALSE)
```

There is enough data. However, we must handle this data in a different way. Instead
of considering the raw numbers we should consider the age of the house and the
number of years since last remodeling. We will create a temporary dataframe to
carry out this analysis.
```{r}
age.data <- training.data[, c("YearBuilt", "YearRemodAdd", "YrSold", "SalePrice")]
age.data$Age <- age.data$YrSold - age.data$YearBuilt
age.data$YrSinceLastMod <- age.data$YrSold - age.data$YearRemodAdd
```

Further, the age of the house and the number of years since last remodeling can
be split into intervals. To do so, we need their summary.
```{r}
summary(age.data$Age)
truehist(age.data$Age, xlab = "Age")
```

We will create levels of $20$ years each for the age of the house.
```{r}
summary(age.data$YrSinceLastMod)
truehist(age.data$YrSinceLastMod, xlab = "# years since last remodeling")
```

An interval of $10$ years might suit the attribute 'years since last modeling'.
We now create the interval attributes for each of these variables.
```{r}
age.data$Age.1 <- cut(age.data$Age, 
                      breaks = seq(from = 0, 
                                   to = round(max(age.data$Age)/10)*10, 
                                   by = 20))
age.data$YrSinceLastMod.1 <- 
    cut(age.data$YrSinceLastMod,
        breaks = seq(from = 0, 
                     to = round(max(age.data$YrSinceLastMod)/10)*10, 
                     by = 10))
```

We will now check if the age and years since last modeling are candidate 
predictors of sale price. We first check the group means and box plots.
```{r}
with(age.data,
     tapply(SalePrice, Age.1, mean))
with(age.data,
     boxplot(SalePrice ~ Age.1, xlab = "Age"))
```

The house price decreases with age up to a limit and then rises as houses grow
older. 

```{r}
with(age.data,
     tapply(SalePrice, YrSinceLastMod.1, mean))
with(age.data,
     boxplot(SalePrice ~ YrSinceLastMod.1, xlab = "Age"))
```

Quite expectedly, recently renovated houses fetch a better price. We will 
confirm our observations about the box plots by ANOVA test.
```{r}
oneway.test(SalePrice ~ Age.1, data = age.data)
oneway.test(SalePrice ~ YrSinceLastMod.1, data = age.data)
rm(age.data)
```

The group means are indeed different.

## Roof style and material
These attributes are nominal data. We first check their availability.
```{r}
data_availability("RoofStyle")
data_availability("RoofStyle")
```

Several roof styles have a single observation. Therefore, RoofStyle might not
be a great choice of an attribute to predict sale price. We now examine the 
group means, ANOVA and box plot of RoofStyle.
```{r}
check_group_means("RoofStyle")
```

The group means of RoofStyle are statistically distinct. We will not be able
to conduct an ANOVA for RoofMatl owing to lack of data. We can compute the
group means, nevertheless.
```{r}
with(training.data, tapply(SalePrice, RoofMatl, length))
with(training.data, tapply(SalePrice, RoofMatl, mean))
```

We may want to combine Membran, WdShake and WdShngl into a single category, 
say 'HighEnd' and the remaining into 'Normal'. Let us see if that makes
a difference.
```{r}
roof.mat.data <- training.data[, c("RoofMatl", "SalePrice")]
roof.mat.data$RoofMatl.1 <- 
    ifelse(roof.mat.data$RoofMatl %in% c("Membran", "WdShake", "WdShngl"), 
           "HighEnd", 
           "Normal")
with(roof.mat.data,
     tapply(SalePrice, RoofMatl.1, mean))
oneway.test(SalePrice ~ RoofMatl.1, data = roof.mat.data)
rm(roof.mat.data)
```

Thus, collapsing levels of roof material into two helps determine the mean sale
price.

## Exterior covering
The two variables Exterior1st and Exterior2nd have $15$ and $16$ levels each.
We first check if we have enough data.
```{r}
data_availability("Exterior1st")
data_availability("Exterior2nd")
```

A few categories of exterior covering have only one or two cases. Therefore, we
cannot run an ANOVA on the eaw data. In order to decide how to combine levels
we look at the group means.
```{r}
X.1 <- with(training.data,
            data.frame(n.samples = tapply(SalePrice, Exterior1st, length),
                       avg.price = tapply(SalePrice, Exterior1st, mean)))
X.1[order(X.1$avg.price), ]
```

We can combine 'AsphShn', 'CBlock' and 'AsbShng' into a single category. We may 
have to leave 'BrkComm' askide because of the significant difference in the 
average price. Let us view the sale prices for 'BrkComm' to make sure that one
of them is not a significant outlier.
```{r}
training.data[training.data$Exterior1st == "BrkComm", "SalePrice"]
```

Both the prices are significantly different from the average price of the next
higher category.

We will now explore possibilities of collapsing levels.

- 'AsphShn', 'CBlock' and 'AsbShng' can be grouped together but we cannot run
  an ANOVA due to lack of data.
- 'MetalSd', 'Wd Sdng' and 'WdShing'
```{r}
oneway.test(SalePrice ~ Exterior1st, 
            data = training.data,
            subset = Exterior1st %in% c("MetalSd", "Wd Sdng", "WdShing"))
``` 

They can ge grouped together.

- 'Stucco' and 'HdBoard' 
```{r}
oneway.test(SalePrice ~ Exterior1st, 
            data = training.data,
            subset = Exterior1st %in% c("Stucco", "HdBoard"))
```

The very large $p$-value suggests that we should not reject the null hypothesis
that the two group means are the same. Thus, we can group them together.

We repeat the exercise with Exterior2nd.
```{r}
X.2 <- with(training.data,
            data.frame(n.samples = tapply(SalePrice, Exterior2nd, length),
                       avg.price = tapply(SalePrice, Exterior2nd, mean)))
X.2[order(X.2$avg.price), ]
```

It would have been nice if we could combine the two summaries. However, the 
attribute names are slightly different. We will, therefore, just join them by 
rows.
```{r}
X.agg <- aggregate(
    cbind(n.samples = SalePrice) ~ Exterior1st + Exterior2nd,
    data = training.data,
    FUN = function(x) {
        NROW(x)
    }
)
X.agg <- X.agg[order(-X.agg$n.samples), ]
X.agg$cum.sum <- cumsum(X.agg$n.samples)
X.agg$cum.pct <- round(X.agg$cum.sum/sum(X.agg$n.samples) * 100, 2)

Y.agg <- aggregate(
    cbind(avg.price = SalePrice) ~ Exterior1st + Exterior2nd,
    data = training.data,
    FUN = function(x) {
        mean(x)
    }
)
Z.agg <- merge(X.agg, Y.agg)
Z.agg <- Z.agg[order(Z.agg$cum.pct), ]
rm(X.agg, Y.agg)
```

We view the aggregate data set.
```{r}
Z.agg
```

It is evident that $95\%$ of the data can be categorized into combinations of
Exterior1st and Exterior2nd which have at least $5$ samples. We, therefore, run
ANOVA on this subset.
```{r}
Z.agg.1 <- Z.agg[Z.agg$n.samples >= 5, ]
Z.data <- merge(training.data, Z.agg.1, by = c("Exterior1st", "Exterior2nd"))
oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.data)
```

The means are not all same. Yet, if we look at the data, we observe that there
is a scope to group some categories together. We will check the same combinations
that we did while investigating Exterior1st alone.

- "MetalSd", "Wd Sdng" and "WdShing"
```{r}
oneway.test(SalePrice ~ Exterior1st + Exterior2nd,
            data = Z.data,
            subset = Exterior1st %in% c("MetalSd", "Wd Sdng", "WdShing"))
```

These levels can be combined together.

- "Stucco" and "HdBoard"
```{r}
oneway.test(SalePrice ~ Exterior1st + Exterior2nd,
            data = Z.data,
            subset = Exterior1st %in% c("Stucco", "HdBoard"))
```

These levels cannot be combined together.

To check if there are more possibilities of combining levels, we will relook at
combinations with at least $5$ samples but now with the data set ordered by 
average sale price.
```{r}
Z.agg.1[order(Z.agg.1$avg.price), c("Exterior1st", "Exterior2nd", "n.samples", "avg.price")]
```

The following pairs of levels can be merged together.

- Combine (WdShing, Plywood) with (Wd Sdng, Wd Shng)
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "WdShing" & Exterior2nd == "Plywood") |
        (Exterior1st == "Wd Sdng" & Exterior2nd == "Wd Shng")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

- (HdBoard, Plywood), (Wd Sdng, Wd Sdng) and (VinylSd, Wd Shng)
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "HdBoard" & Exterior2nd == "Plywood") |
        (Exterior1st == "Wd Sdng" & Exterior2nd == "Wd Sdng") |
        (Exterior1st == "VinylSd" & Exterior2nd == "Wd Shng")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

- (Plywood, Brk Cmn), (MetalSd, MetalSd) and (Wd Sdng, Plywood)
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "Plywood" & Exterior2nd == "Brk Cmn") |
        (Exterior1st == "MetalSd" & Exterior2nd == "MetalSd") |
        (Exterior1st == "Wd Sdng" & Exterior2nd == "Plywood")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

- (Stucco, Stucco), (WdShing, Wd Shng) and (HdBoard, HdBoard)
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "Stucco" & Exterior2nd == "Stucco") |
        (Exterior1st == "WdShing" & Exterior2nd == "Wd Shng") |
        (Exterior1st == "HdBoard" & Exterior2nd == "HdBoard")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

They can be combined together.

- (BrkFace, Wd Sdng), (BrkFace,	Plywood) and (BrkFace, BrkFace)
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "BrkFace" & Exterior2nd == "Wd Sdng") |
        (Exterior1st == "BrkFace" & Exterior2nd == "Plywood") |
        (Exterior1st == "BrkFace" & Exterior2nd == "BrkFace")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

They can be combined together.

- (VinylSd, VinylSd) and (CemntBd, CmentBd). Notice the slight difference in 
  spelling in the second pair.
```{r}
Z.subset <- subset(
    Z.data,
    subset = 
        (Exterior1st == "VinylSd" & Exterior2nd == "VinylSd") |
        (Exterior1st == "CemntBd" & Exterior2nd == "CmentBd")
)

oneway.test(SalePrice ~ Exterior1st + Exterior2nd, data = Z.subset)
```

I think, these variables are hard to use. If at all they are needed to build a
model, we must

- restrict ourselves to those combinations of the two for which there are at 
  least $5$ samples.
- explore possibilities of combining data.

```{r}
rm(X.1, X.2, Z.agg, Z.agg.1, Z.data, Z.subset)
```

## Masonry veneer 
We start with checking data.
```{r}
data_availability("MasVnrType")
```

A related variable is MasVnrArea. Let us explore their relationship through
box plots.
```{r}
with(training.data,
     boxplot(MasVnrArea ~ MasVnrType))
```

It is quite strange to see the category 'None' having non-zero area and the
other categories having a zero area. We will summarize the area for each
category.
```{r}
show_summary <- function(vnr.type) {
    with(training.data[training.data$MasVnrType == vnr.type, ], summary(MasVnrArea))
}

sapply(levels(training.data$MasVnrType), show_summary)
```

Before exploring the relationship of veneer type, veneer area and sale price, we
will look at the group means of veneer type alone.
```{r}
check_group_means("MasVnrType")
```

Is having 'BrkCmn' no different from having 'None'?
```{r}
oneway.test(SalePrice ~ MasVnrType,
            data = training.data,
            subset = MasVnrType %in% c("BrkCmn", "None"))
```

Looks like they are no different and can be combined in a single category.
Let us see how sale price varies with veneer area for all types except 'None'.

```{r}
xyplot(SalePrice ~ MasVnrArea | MasVnrType, data = training.data, cex = 0.1)
```

As expected, there is a positive correlation between veneer area and sale price.
Let us find the correlation coefficients for each veneer type.
```{r}
tmp.data <- training.data[!is.na(training.data$MasVnrArea), ]
by(tmp.data[, c("MasVnrArea", "SalePrice")], 
   tmp.data$MasVnrType, 
   FUN = function(x) {cor(x$MasVnrArea, x$SalePrice)})
rm(tmp.data)
```

There is a reasonably strong correlation between masonry veneer area and sale
price when the masonry veneer type is not 'None'.

## External quality and condition
We expect these variables to be similar to overall quality and condition. 
```{r}
data_availability("ExterQual")
data_availability("ExterCond")
```

The group means, ANOVA and box plot for levels of external quality are
```{r}
check_group_means("ExterQual")
```

External quality strongly detemines the sale price. We cannot repeat this
exercise for external condition because there is only case when external 
condition is poor.
```{r}
check_group_means("ExterCond", run.anova = FALSE)
```

We also expect these variables to be strongly correlated.
```{r}
table(training.data[, c("ExterQual", "ExterCond")])
```

External condition is also a strong predictor of sale price.

## Basement quality and condition
Their data availability is
```{r}
data_availability("BsmtQual")
data_availability("BsmtCond")
```

Their group means are
```{r}
check_group_means("BsmtQual")
check_group_means("BsmtCond")
```

Like the other quality and condition variables, these ones too strongly decide
the sale price.

## Basement exposure
```{r}
data_availability("BsmtExposure")
```

The group means analysis is
```{r}
check_group_means("BsmtExposure")
```

## Basement finishing
BsmtFinType1 and BsmtFinType2 describe the conditions of the basement fittings.
```{r}
data_availability("BsmtFinType1")
data_availability("BsmtFinType2")
```

An analysis of their group means is
```{r}
check_group_means("BsmtFinType1")
check_group_means("BsmtFinType2")
```

Some group means are quite close to each other.

- BLQ, LwQ and Rec can be grouped together.
```{r}
oneway.test(SalePrice ~ BsmtFinType1,
            data = training.data,
            subset = BsmtFinType1 %in% c('BLQ', 'LwQ', 'Rec'))
oneway.test(SalePrice ~ BsmtFinType2,
            data = training.data,
            subset = BsmtFinType2 %in% c('BLQ', 'LwQ', 'Rec'))
```

- ALQ and Unf. Although the $p$-value is greater than the customary $5\%$, I am
inclined to keep them separate.
```{r}
oneway.test(SalePrice ~ BsmtFinType1,
            data = training.data,
            subset = BsmtFinType1 %in% c('ALQ', 'Unf'))
```

## Basement area
There are four numeric variables BsmtFinSF1, BsmtFinSF2, BsmtUnfSF, TotalBsmtSF.
We will check their availability before looking at the trends of sale price.
```{r}
data_availability("BsmtFinSF1", show.grp.size = FALSE)
data_availability("BsmtFinSF2", show.grp.size = FALSE)
data_availability("BsmtUnfSF", show.grp.size = FALSE)
data_availability("TotalBsmtSF", show.grp.size = FALSE)
```

Variation of sale price for BsmtFinSF1 is
```{r}
with(training.data[training.data$BsmtFinSF1 != 0, ], 
     plot(BsmtFinSF1, SalePrice, cex = 0.1))
```

Basement area of type 1 influences sale price.

```{r}
with(training.data[training.data$BsmtFinSF2 != 0, ], 
     plot(BsmtFinSF2, SalePrice, cex = 0.1))
```

Basement area of type 2 is not a very strong influencer of sale price.

```{r}
with(training.data[training.data$BsmtUnfSF != 0, ], 
     plot(BsmtUnfSF, SalePrice, cex = 0.1))
```

Unfinished area of the basement is not a very strong influencer of sale price.

```{r}
with(training.data[training.data$TotalBsmtSF != 0, ], plot(TotalBsmtSF, SalePrice))
```

Sale price is strongly correlated to total basement area. It might help to look
at the variance-covariance matrix of these four variables and sale price.
```{r}
cor(training.data[, c("BsmtFinSF1",
                      "BsmtFinSF2",
                      "BsmtUnfSF",
                      "TotalBsmtSF",
                      "SalePrice")])
```

Only the total basement area is strongly correlated with sale price.

## Heating 
The variables 'Heating' and 'HeatingQC' are categorical variables. We carry out
the usual analysis.
```{r}
data_availability("Heating")
data_availability("HeatingQC")
```

The data availability is too skewed in the case of 'Heating'. The group means,
without ANOVA are
```{r}
check_group_means("Heating", run.anova = FALSE)
```

How close are 'GasA' and 'GasW'?
```{r}
oneway.test(SalePrice ~ Heating,
            data = training.data,
            subset = Heating %in% c("GasA", "GasW"))
```

Perhaps they can be grouped together. But that might not be of much use as there
are very few cases for other levels.

Moving to heating quality and condition, 
```{r}
check_group_means("HeatingQC", run.anova = FALSE)
```

There is only case of poor condition of heating. Let us conduct an ANOVA for the
remaining data.
```{r}
oneway.test(SalePrice ~ HeatingQC,
            data = training.data,
            subset = HeatingQC != "Po")
```

The group means are distinct. The sale price of the case with poor quality 
heating is also quite apart from the remaining means.

## Central air-conditioning
```{r}
data_availability("CentralAir")
```

```{r}
check_group_means("CentralAir")
```

## Electrical
```{r}
data_availability("Electrical")
```

We will check the group means without ANOVA because some levels have few cases.
```{r}
check_group_means("Electrical", run.anova = FALSE)
```

The group means are sufficiently distinct to do away with ANOVA. Yet, we 
carry it out to support our belief.
```{r}
oneway.test(SalePrice ~ Electrical,
            data = training.data,
            subset = Electrical != "Mix")
```

## Floor area
There are four numeric variables, X1stFlrSF, X2ndFlrSF, LowQualFinSF and 
GrLivArea related to area of the house. We will check their availability before 
looking at the trends of sale price.
```{r}
data_availability("X1stFlrSF", show.grp.size = FALSE)
data_availability("X2ndFlrSF", show.grp.size = FALSE)
data_availability("LowQualFinSF", show.grp.size = FALSE)
data_availability("GrLivArea", show.grp.size = FALSE)
```

```{r}
with(training.data, plot(X1stFlrSF, SalePrice))
with(training.data[training.data$X2ndFlrSF != 0, ], plot(X2ndFlrSF, SalePrice, cex = 0.1))
with(training.data[training.data$LowQualFinSF != 0, ], plot(LowQualFinSF, SalePrice, cex = 0.1))
with(training.data, plot(GrLivArea, SalePrice, cex = 0.1))
```

Not surprisingly, the above grade (ground) living area square feet is strongly
correlated with sale price. Before moving on, we confirm that GrLivArea is the
sum of the remaining three.
```{r}
with(training.data, any(GrLivArea - X1stFlrSF - X2ndFlrSF - LowQualFinSF != 0))
```

## Bathrooms
The number of bathrooms is going to be limited. Therefore, when we check 
availability of data, we can display the group data.
```{r}
data_availability("BsmtFullBath")
data_availability("BsmtHalfBath")
data_availability("FullBath")
data_availability("HalfBath")
```

We now look at the group means.
```{r}
check_group_means("BsmtFullBath", run.anova = FALSE)
check_group_means("BsmtHalfBath")
check_group_means("FullBath")
check_group_means("HalfBath")
```

The number of half-basements in the basement do not matter to the sale price. 
The number of full bathrooms in the house do matter, so do the number of half
bathrooms in the house. But of the two, the former variable seems to be showing
the greatest impact.

## Bedrooms above ground
```{r}
data_availability("BedroomAbvGr")
```

There is only case with $8$ bedrooms. Therefore, we will not run a full ANOVA.
```{r}
check_group_means("BedroomAbvGr", run.anova = FALSE)
```

I would have expected sale price to be a monotone function of the number of
bedrooms. Therefore, it is baffling that the group mean of houses with zero 
bedrooms is the highest. Let us investigate the group more.
```{r}
with(training.data[training.data$BedroomAbvGr == 0, ], summary(SalePrice))
```

The range of prices is quite large. Moreover, the sample is quite small.
```{r}
training.data[training.data$BedroomAbvGr == 0, c("SalePrice")]
```

Although the house prices are 'well-resolved' by the number of bedrooms above
ground, the trend is counter-intuitive. Therefore, I would hesitate building a
model using it.

## Kitchen
```{r}
data_availability("KitchenAbvGr")
data_availability("KitchenQual")
```

All except three houses have one or two kitchens. I will, therefore, run an
ANOVA only for these two categories.
```{r}
oneway.test(SalePrice ~ KitchenAbvGr,
            data = training.data,
            subset = KitchenAbvGr %in% c(1, 2))
```

The number of Kitchens, if they are either $1$ or $2$, do result in different
sale price. The group means are
```{r}
aggregate(cbind(avg.price = SalePrice) ~ KitchenAbvGr, 
          data = training.data, 
          FUN = function(x) mean(x))
```

Going by past experience, I think that the quality of the kitchen will be a 
significant predictor of sale price than the number of kitchens.
```{r}
check_group_means("KitchenQual")
```

The quality of the kitchen indeed matters more.

## Total number of rooms above ground.
Let us find if the total number of rooms has the same behavior as the total 
number of bedrooms.
```{r}
data_availability("TotRmsAbvGrd")
```

The group means are
```{r}
check_group_means("TotRmsAbvGrd", run.anova = FALSE)
```

The behavior is not as baffling as in the number of bedrooms above ground. The
average price rises monotonically up to $11$ rooms. It drops thereafter.

## Functionality
The meaning of this variable is not very clear to me.
```{r}
data_availability("Functional")
```

The group means are
```{r}
check_group_means("Functional", run.anova = FALSE)
```

We will look at the ANOVA results by dropping the levels with only one case.
```{r}
oneway.test(SalePrice ~ Functional,
            data = training.data,
            subset = Functional != "Sev")
```

The differences in groups means are statistically significant.

## Fireplaces
```{r}
data_availability("Fireplaces")
data_availability("FireplaceQu")
```

```{r}
check_group_means("Fireplaces")
check_group_means("FireplaceQu")
```

The group means of the quality variable are more apart from each other than just
the number.

## Garage
```{r}
data_availability("GarageType")
```

```{r}
check_group_means("GarageType")
```

Are garage types '2Types' and 'Basment' sufficiently different?
```{r}
oneway.test(SalePrice ~ GarageType,
            data = training.data,
            subset = GarageType %in% c("2Types", "Basment"))
```

They are not.

The year in which a garage was built can be used to determine its weight. We will
therefore find out how old the garage was when the house was sold. But before that,
we will check if there are NAs.
```{r}
data_availability("GarageYrBlt", show.grp.size = FALSE)
```

```{r}
X.1 <- training.data[!is.na(training.data$GarageYrBlt), c("GarageYrBlt", "YrSold", "SalePrice")]
X.1$garage.age <- X.1$YrSold - X.1$GarageYrBlt
```

Summary of garage age is
```{r}
summary(X.1$garage.age)
```

Let us view the histogram of age of the garage before splitting them into 
intervals.
```{r}
truehist(X.1$garage.age, xlab = "Age of the garage")
```

An interval of $10$ years seems appropriate.
```{r}
X.1$garage.age.1 <- cut(X.1$garage.age,
                        seq(from = 0, 
                            to = round(max(X.1$garage.age)/10)*10, 
                            by = 10))
```

We will now examine the group means.
```{r}
with(X.1, tapply(SalePrice, garage.age.1, mean))
oneway.test(SalePrice ~ garage.age.1, data = X.1)
```

Although the group means are not all identical, neither are they monotonic. I
suspect that the age of the garage has only a weak influence on sale price.

```{r}
rm(X.1)
```

Let us now consider the garage finish.
```{r}
data_availability("GarageFinish")
```

There are a few NAs. We remove them.
```{r}
X.1 <- training.data[!is.na(training.data$GarageFinish), ]
with(X.1, tapply(SalePrice, GarageFinish, mean))
oneway.test(SalePrice ~ GarageFinish, data = X.1)
rm(X.1)
```

Garage finish seems to be good predictor of the sale price.

Moving on to the number of cars that can be parked in the garage.
```{r}
data_availability("GarageCars")
```

There are enough cases in each level. We can, therefore, look at the group means.
```{r}
check_group_means("GarageCars")
```

The sale price is well-resolved though higher capacity does not give higher sale
price. To get th reasons, we look at the five cases with a garage capacity of 
four cars.
```{r}
training.data[training.data$GarageCars == 4, c("SalePrice")]
```

One of the houses was sold very cheap.

We now consider the quality of the garage.
```{r}
data_availability("GarageQual")
```

Although the data is skewed, we can still consider group means.
```{r}
check_group_means("GarageQual")
```

The quality of the garage is a strong predictor of the house price.

We now conside the last variable related to garage, its condition.
```{r}
data_availability("GarageCond")
```

Once again the data is skewed yet there are enough cases to carry out ANOVA.
```{r}
check_group_means("GarageCond")
```

The levels 'Fa' and 'Po' seem to have means close to each other. Let us examine
them separately.
```{r}
oneway.test(SalePrice ~ GarageCond,
            data = training.data,
            subset = GarageCond %in% c("Fa", "Po"))
```

The differences in the group means are not statistically significant.

## Paved drive
```{r}
data_availability("PavedDrive")
```

The group means analysis is
```{r}
check_group_means("PavedDrive")
```

Are the group means of levels N and P statistically different?
```{r}
oneway.test(SalePrice ~ PavedDrive,
            data = training.data,
            subset = PavedDrive %in% c("N", "P"))
```

The differences in group means are statistically significant.

## Wooddeck 
It is a numeric variable describing the area in square feet of the wooden deck.
```{r}
data_availability("WoodDeckSF", show.grp.size = FALSE)
```

A summary of the data is
```{r}
summary(training.data$WoodDeckSF)
```

If the median is zero then more than half of the houses do not have a wooden
deck. When it is present, does it have an impact on the sale price?
```{r}
xyplot(SalePrice ~ WoodDeckSF, 
       data = training.data[training.data$WoodDeckSF > 0, ],
       cex = 0.3)
```

The correlation does not appear to be very strong. In any case, if this varible
is significant for less than half of the cases, it might not be very useful in
predicting the sale price.

## Porch
OpenPorchSF is a numeric variable describing the area in square feet of the porch.
```{r}
data_availability("OpenPorchSF", show.grp.size = FALSE)
```

A summary of the data is
```{r}
summary(training.data$OpenPorchSF)
```

At least a quarter of the cases do not have an open porch. Nevertheless, for 
those that have it, let us look at the trend of sale price versus porch area.
```{r}
xyplot(SalePrice ~ OpenPorchSF,
       data = training.data[training.data$OpenPorchSF != 0, ],
       cex = 0.3)
```

The correlation does not appear to be very strong.

EnclosedPorch is a numeric variable describing measuring the area of enclosed
porch.
```{r}
data_availability("EnclosedPorch", show.grp.size = FALSE)
```

A summary of the data is
```{r}
summary(training.data$EnclosedPorch)
```

A very large number of houses do not have an enclosed porch. 
```{r}
nrow(training.data[training.data$EnclosedPorch == 0, ])
```

Moving to X3SsnPorch, the area of three seasons porch. Its summary is
```{r}
summary(training.data$X3SsnPorch)
```

More than $75\%$ of the houses do not have it.

The last porch related variable is ScreenPorch. It is the area of screened porch.
Its summary is
```{r}
summary(training.data$ScreenPorch)
```

More than $75\%$ of the houses do not have it.

## Pool
A large number of houses do not have a pool. A summary of PoolArea variable 
should confirm this understanding.
```{r}
summary(training.data$PoolArea)
```

Nevertheless, we look at the sale price as a function of pool condition, whenever
it is present.
```{r}
with(training.data[!is.na(training.data$PoolQC), ],
     tapply(SalePrice, PoolQC, length))
with(training.data[!is.na(training.data$PoolQC), ],
     tapply(SalePrice, PoolQC, mean))
```

Only $7$ houses have a data about pool quality.

## Fence
```{r}
data_availability("Fence")
```

An analysis of the group means is
```{r}
check_group_means("Fence")
```

Let us check if the categories 'GdWo' and 'MnWw' are distinct.
```{r}
oneway.test(SalePrice ~ Fence,
            data = training.data,
            subset = Fence %in% c("GdWo", "MnWw"))
```

They are not. What about 'GdWo' and 'MnPrv'?
```{r}
oneway.test(SalePrice ~ Fence,
            data = training.data,
            subset = Fence %in% c("GdWo", "MnPrv"))
```

They too are not distinct.

## Miscellaneous features
```{r}
data_availability("MiscFeature")
```

Only $3.7\%$ cases had a miscellaneous feature. Therefore, we may not find this
attribute useful.
