---
title: "Sampling Distributions"
author: "Amey Joshi"
date: "16/11/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
set.seed(18081932)
```

## The idea

If we draw $n$ samples from a population and compute the mean for each one of 
them we find that the means are not identical. The distribution of sample means
is called the sampling distribution of means. We can extend this idea to any
statistic. We demonstrate how to study sampling distribution of means and 
variances of water level of Lake Huron, one of the five great lakes in the 
north America.

The data set of water level of Lake Huron is available on every R installation.

## Visual examination of the data
A summary of the data is
```{r}
data.summary <- summary(LakeHuron)
print(data.summary)
```

The water level has a relatively narrow range about the mean.
```{r, echo = TRUE}
upper.fluctuation <- (data.summary[6] - data.summary[4])/data.summary[4]
lower.fluctuation <- (data.summary[4] - data.summary[1])/data.summary[4]

up <- round(upper.fluctuation * 100, 2)
cat(paste(sep = "", "The highest level is ", up, "% above the mean.\n"))

dn <- round(lower.fluctuation * 100, 2)
cat(paste(sep = "", "The lowest level is ", dn, "% below the mean.\n"))
```

The highest and the lowest levels of the lake are not symmetrically placed about
the mean. We expect the distribution of water level to be asymmetric. We check 
it by plotting the histogram with its empirical density.
```{r, echo = TRUE}
truehist(LakeHuron, nbins = 14, main = "Water level of Lake Huron")
lines(density(LakeHuron))
rug(LakeHuron)
```

## Is the data normal?
In order to study the sampling distribution we first examine if the data is normal.
We use the Shapiro-Wilk test whose null hypothesis is that the data is indeed
normal.
```{r, echo = TRUE}
sig.p <- 0.05 # Significant p-value.
normality.result <- shapiro.test(LakeHuron)
p.val <- round(normality.result[["p.value"]], 3)
if (p.val > sig.p) {
  cat(paste(sep = "", "Shapiro-Wilk test has p-value ", p.val, ".\n"))
  cat(paste(sep = "", "Data is normal ", "at significance level ", sig.p, ".\n"))
} else {
  cat(paste(sep = "", "Shapiro-Wilk test has p-value ", p.val, ".\n"))
  cat(paste(sep = "", "Data is NOT normal ", "at significance level ", sig.p, ".\n"))
}
```

## Sampling distribution of means
If the population data is normally distributed then the standardized sample means
are $t$-distributed with $\nu = n - 1$ degrees of freedom, $n$ being the sample
size.
```{r, echo = TRUE}
population.size <- length(LakeHuron)
sample.size <- population.size / 7
n.samples <- 500

sample_stats <- function(i) {
  # This function has a dummy argument. We have it so that it can be invoked
  # through an apply family of functions.
  s <-
    LakeHuron[sample(population.size, sample.size, replace = TRUE)]
  c(mean(s), sd(s))
}

sample.stats <- sapply(1:n.samples, sample_stats)
population.mean <- mean(LakeHuron)
sample.t <-
  (sample.stats[1,] - population.mean) / (sample.stats[2,] / sqrt(sample.size))

# Get the domain of the t-distribution that best fits our standardized sample 
# means.
L <- round(max(abs(range(sample.t))))
X <- seq(from = -L, to = L, length.out = n.samples)
Y <- dt(X, df = sample.size - 1)
plot(X, Y, type = "l", lty = 1, col = 1)
lines(density(sample.t), lty = 2, col = 2)
legend.text <- c("Theoretical", "Actual")
legend("topright", col = c(1, 2), lty = c(1, 2), legend = legend.text, cex = 0.8)
title(main = "Sampling distribution of means")
```

We observe the distribution of sample means is close to the theoretically 
expected $t$ distribution with $6$ degrees of freedom.

## Sampling distribution of variances
If the population data is normally distributed and the sample size if $n$ then
we compute $(n - 1)s^2/\sigma^2$, where $s^2$ is the sample variance and 
$\sigma^2$ is the population variance. This quantity has as $\chi^2$-distribution
with $n - 1$ degrees of freedom. 
```{r, echo = TRUE}
sample.var <- sample.stats[2, ]^2

# The var function of R always computes the sample variance.
population.var <- var(LakeHuron) * (sample.size - 1)/sample.size

sample.chisq <- (sample.size - 1) * sample.var/population.var
X.left <- seq(from = 0, to = 10, length.out = 100)
X.right <- seq(from = 10, to = 30, length.out = n.samples - length(X.left))
X <- c(X.left, X.right)
Y <- dchisq(X, df = sample.size - 1)
# We compute the density separately so that we can estimate the extent of the
# y-axis.
D <- density(sample.chisq)
plot(X, Y, type = "l", lty = 1, col = 1, ylim = c(0, round(range(D$y)[2], 1)))
lines(D, lty = 2, col = 2)
legend.text <- c("Theoretical", "Actual")
legend("topright", col = c(1, 2), lty = c(1, 2), legend = legend.text, cex = 0.8)
title(main = "Sampling distribution of variances")
```

The sampling distribution of variances is approximately the same as the theorerically
expected $\chi^2$-distribution with $6$ degrees of freedom.
