---
title: "Dummy Variables"
author: "Amey Joshi"
format: html
  df-print: kable
editor: visual
---

# Multiple regression

## Examination of the data

Let us once again consider the `swiss` dataset and examine the dependent variables in it.

```{r}
library(datasets)
data("swiss")
head(swiss)
```

A histogram of the variable `Catholic` is:

```{r}
hist(swiss$Catholic, xlab='% Catholics', ylab='Frequency', main='Catholics in Swiss cantons')
```

Its density looks like:

```{r}
plot(density(swiss$Catholic), main='Catholics in Swiss cantons')
```

The density is bimodal and it may perhaps not be a bad idea to create a binary variable which it $1$ in cantons with more than $50\%$ Catholics and $0$ in the rest.

## Creation of a binary variable

```{r}
swiss$Catholic.Bin <- 1 * (swiss$Catholic > 50)
head(swiss)
```

Let us check how many cantons have a Catholic majority.

```{r}
aggregate(Fertility ~ Catholic.Bin, data=swiss, length)
```

We will check the variation of `Fertility` with `Agriculture` depending on whether the Catholics have a majority in the canton.

```{r}
plot(swiss$Agriculture,
     swiss$Fertility,
     col = ifelse(swiss$Catholic.Bin == 1, 'blue', 'red'),
     pch = 16,
     xlab = 'Agriculture',
     ylab = 'Fertility',
     main = 'Swiss fertility by Catholic population')
legend(x = 'bottomright',
       legend = c('Majority', 'Minority'),
       pch = c(16, 16),
       col = c('blue', 'red'),
       bty = 'n',
       cex = 1)
```

The Catholic majority cantons have higher percentage of males practising agriculture and a high fertility as well.

## The model

Let $X_1$ denote `Agriculture` and the categorical variable $X_2$ denote `Catholic.Bin`. We will try to fit the model $$
Y_i = \beta_0 + \beta_1 X_{1,i} + \beta_2 X_{2, i} + \epsilon_i.
$$ Since $X_2$ takes only two values, $0$ and $1$, we really are fitting the lines $$
\begin{eqnarray}
Y_i &=& \beta_0 + \beta_1 X_{1,i} + \beta_2 + \epsilon_i \\
    &=& (\beta_0 + \beta_2) + \beta_1 X_{1, i} + \epsilon_i \\
\end{eqnarray}
$$ for Catholic majority cantons and $$
Y_i = \beta_0 + \beta_1 X_{1, i} + \epsilon_i
$$ for cantons in which the Catholics are a minority. So we are really fitting two lines with same slope but differing intercepts. If the dummy variable had $N$ levels then we would have a set of $N$ parallel lines.

```{r}
res.1 <- lm(Fertility ~ Agriculture + Catholic.Bin, data=swiss)
summary(res.1)
```

The coefficient of `Catholic.Bin` is the change in intercept when we go from a Catholic minority canton to a Catholic majority one. One can choose to experiment with a model where the both the slope and the intercept change among the two categories of cantons.

## Interaction terms

A model with interaction has the form $$
Y_i = \beta_0 + \beta_1 X_{1,i} + \beta_2 X_{2, i} + \beta_3 X_{1, i}X_{2, i}\epsilon_i.
$$ When $X_2 = 0$ then it reduces to $Y_i = \beta_0 + \beta_1 X_{1, i} + \epsilon_i$. However, when $X_2 = 1$ it becomes $$
Y_i = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X_{1, i} + \epsilon_i.
$$ This line has different slope and intercept. Let us fit such a model.

```{r}
res.2 <- lm(Fertility ~ Agriculture * Catholic.Bin, data=swiss)
summary(res.2)

```

Let us ignore the $p$-values of the coefficients once again. The intercept increases by an amount equal to the coefficient of `Catholic.Bin` when `Catholic.Bin = 1`. Simultaneously, the slope increases by an amount equal to the coefficient of `Agriculture:Catholic.Bin`.

```{r}
coeffs <- res.2$coefficients
with(swiss,
     plot(Agriculture,
          Fertility,
          col = ifelse(Catholic.Bin == 1, 'blue', 'red'),
          xlab = 'Agriculture',
          ylab = 'Fertility',
          main = 'Swiss fertility by Catholic population'))
abline(coeffs[1], coeffs[2], col = 'red')
abline(coeffs[1] + coeffs[3], coeffs[2] + coeffs[4], col='blue')
legend(x = 'bottomright',
       legend = c('Majority', 'Minority'),
       pch = c(16, 16),
       col = c('blue', 'red'),
       bty = 'n',
       cex = 1)
```
