---
title: "Simulations"
author: "Amey Joshi"
format: html
editor: visual
---

## Subtleties in analysis

The relationship between the dependent variable $Y$ and the independent variables $X_i$ is often influenced by other variables not included in $\{X_i\}$. Often times, these omitted variables are categorical. We will consider examples of such data by simulating it.

## Simulation 1

We simulate the data as:

```{r}
n <- 100 # number of observations.
t <- rep(c(0, 1), c(n/2, n/2)) # A categorical variable.
x <- c(runif(n/2), runif(n/2)) 
beta0 <- 0
beta1 <- 2
tau <- 1
sigma <- 0.2

y <- beta0 + beta1 * x + tau * t + rnorm(n, 0, sigma)
```

The data looks like

```{r}
plot(x, y, col = ifelse(t == 1, 'red', 'blue'), pch = 16)
abline(a = beta0, b = beta1, col = 'black', lwd = 3)
abline(a = beta0 + tau, b = beta1, col = 'black', lwd = 3)
title('Grouped data')
legend(x = 'topleft',
       legend = c('Treatment', 'No treatment'),
       pch = c(16, 16),
       col = c('red', 'blue'),
       bty = 'n',
       cex = 1)

```

### Model with treatment variable ignored

Supposing we ignore the presence of the treatment variable `G` and fit a linear model.

```{r}
res.1 <- lm(y ~ x)
agg.res <- aggregate(y ~ t, FUN = mean)
summary(res.1)

plot(x, y, col = ifelse(t == 1, 'red', 'blue'), pch = 16)
abline(res.1$coefficients, col = 'black', lty = 2)
abline(h = agg.res[agg.res$t == 0, 'y'], col = 'blue')
abline(h = agg.res[agg.res$t == 1, 'y'], col = 'red')
legend(x = 'topleft',
       legend = c('Treatment', 'No treatment'),
       pch = c(16, 16),
       col= c('red', 'blue'),
       bty = 'n',
       cex = 1)
title('Grouped data')
```

The model certainly fits well, but it is clearly visible that there is a categorical variable that neatly divides the population in two groups. A few more observations:

1.  The outcome $y$ is depends on the predictor $x$ but the relationship is determined by the group variable $t$.
2.  The outcome $y$ is influenced by $t$ but the predictor $x$ is not.
3.  The dashed line will be the regression line with $t$ ignored.

We will now fit separate model for each group.

```{r}
res.grp.1 <- lm(y ~ x, data = df.1, subset = (df.1$t == 1))
res.grp.0 <- lm(y ~ x, data = df.1, subset = (df.1$t == 0))

plot(x, y, col = ifelse(t == 1, 'red', 'blue'), pch = 16)
abline(res.1$coefficients, col = 'black', lty = 2)
abline(res.grp.1$coefficients, col = 'red', lty = 2)
abline(res.grp.0$coefficients, col = 'blue', lty = 2)
abline(h = agg.res[agg.res$t == 0, 'y'], col = 'blue')
abline(h = agg.res[agg.res$t == 1, 'y'], col = 'red')
legend(x = 'topleft',
       legend = c('Treatment', 'No treatment'),
       pch = c(16, 16),
       col= c('red', 'blue'),
       bty = 'n',
       cex = 1)
title('Grouped data')
```

If ignore $x$, our prediction of $y$ will be the group means. By including $x$, we make them more accurate by reading off the two regression lines. The difference between the regression lines and the corresponding intercepts is surprisingly small. This is because the $x$ variable is evenly distributed across the groups. In such situations, when knowing $x$ does not drastically change our estimate of $y$ the remaining variable $x$ is called a *nuisance variable*.

# Simulation 2

In the previous simulation, the categorical variable resulted in two groups having same slope but different intercepts. Now we consider a situation when both the slope and the intercept are the same but the range of predictor and response variables are different for the two groups.

```{r}
n.sim <- 100
t <- c(rep(0, n.sim/2), rep(1, n.sim/2))
x <- c(runif(n.sim/2), 1.5 + runif(n.sim/2))
beta0 <- 0
beta1 <- 2
tau <- 0
sigma <- 0.2

y = beta0 + beta1 * x + tau * t + rnorm(n.sim, sd = sigma)
plot(x, y, col = ifelse(t == 1, 'blue', 'red'), pch = 16)
legend(x = 'bottomright',
       legend = c('Treatment', 'No treatment'),
       pch = c(16, 16),
       col = c('blue', 'red'),
       bty = 'n',
       cex = 1)
title(main = 'Simulation 2')

full.res <- lm(y ~ x)
trt.res  <- lm(y[1:n.sim/2] ~ x[1:n.sim/2])
ntrt.res <- lm(y[n.sim/2 + 1: n.sim] ~ x[n.sim/2 + 1: n.sim])

abline(full.res$coefficients, col = 'black', lty = 5)
abline(trt.res$coefficients, col = 'red', lty = 5)
abline(ntrt.res$coefficients, col = 'blue', lty = 5)
```

The long-dashed red line fits only the 'No treatment' group, the long-dashed blue line only the 'treatment' group and the long-dashed black line the entire data set. We first compare this data set with the previous one.

1.  In 'simulation 1', $x$ did not depend on $t$, it does in this case.
    1.  If you know $x$ in 'simulation 2', you can very accurately tell the value of $t$, but not *vice versa*.
2.  In both cases, knowing $t$ gives a reasonable estimate of $y$. It is the group mean in both cases.
3.  In 'simulation 1',i ncluding $x$ in the model only improves the accuracy. In this case, including $x$ and $t$ erases, confounds, hides the influence of $t$.

## Simulation 3

The categorical variable $t$ separates the $x$ for the two groups as in the previous case but does not get fully confounded by $x$.

```{r}
n.sim <- 100
t <- c(rep(0, n.sim/2), rep(1, n.sim/2))
x <- c(runif(n.sim/2), 0.9 + runif(n.sim/2))
beta0 <- 0
beta1 <- 2
tau <- -1
sigma <- 0.2

grp.means <- aggregate(y ~ t, FUN = mean)

y = beta0 + beta1 * x + tau * t + rnorm(n.sim, sd = sigma)
plot(x, y, col = ifelse(t == 1, 'blue', 'red'), pch = 16)
legend(x = 'bottomright',
       legend = c('Treatment', 'No treatment'),
       pch = c(16, 16),
       col = c('blue', 'red'),
       bty = 'n',
       cex = 1)
title(main = 'Simulation 3')

full.res <- lm(y ~ x + t)
trt.res  <- lm(y[1:n.sim/2] ~ x[1:n.sim/2])
ntrt.res <- lm(y[n.sim/2 + 1: n.sim] ~ x[n.sim/2 + 1: n.sim])

#abline(full.res$coefficients, col = 'black', lty = 5)
abline(trt.res$coefficients, col = 'red', lty = 5)
abline(ntrt.res$coefficients, col = 'blue', lty = 5)
abline(h = grp.means[grp.means$t == 0, 'y'], lty = 3, col = 'red')
abline(h = grp.means[grp.means$t == 1, 'y'], lty = 3, col = 'blue')
```

The chart has the following lines:

1.  The long-dashed, black line is the fit of the combined data.
2.  The long-dashed red and blue lines are fits of grouped data.
3.  The dotted red and blue lines are the grouped means of $y$. They are our estimates of $y$ if we ignore $x$.

Let us now compare the results of the two models.

### Ignore $x$

```{r}
summary(lm(y ~ t))
```

Here $y$ is positively correlated with $t$. The treatment group has higher estimate than the control group.

### Use both variables

```{r}
summary(lm(y ~ t + x))
```

Here the coefficient of $t$ is negative! The treatment appears to have an effect contrary to the previous model.
