---
title: "Advertisement of a Casino"
author: "Amey Joshi"
date: "2023-03-06"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Betting Mechanism

- A casino has only one game. It is played using a perfect, absolutely unbiased
coin. Each time you toss a coin and it lands head, you earn a rupee. Otherwise,
you lose a rupee.

- There is no charge taken upfront. Each person tosses $500$ times. At the end
he collects cash if he wins or pays money if he loses.

- The casino wants to attract more players. It publishes advertisements in the
print and electronic media about the average profit earned by people who played.

## An example of the game

```{r, echo=FALSE}
set.seed(12111842)
n_trials <- 500
outcomes <- cumsum(sample(c(-1, 1), n_trials, TRUE))
plot(
  seq(n_trials),
  outcomes,
  type = 'l',
  xlab = 'trial',
  ylab = 'payoff',
  main = 'A sample game'
)
payoff <- outcomes[n_trials]
cat('The person earned', payoff, 'rupees.', sep=' ')
```

## Data for the advertisement

```{r cars, echo = TRUE}
n_trials <- 500
n_people <- 1000

gamble <- function(p) {
  cumsum(sample(x=c(-1, 1), size=n_trials, replace=TRUE))[n_trials]
}

payoffs <- unlist(lapply(seq(n_people), gamble), recursive = FALSE)
profits <- payoffs[payoffs > 0]

cat('Average profit was ', round(mean(profits), 2), ' rupees.', sep=' ')
```

## The advertisement

- You pay nothing to start a game.
- Players' average profit is $18.9$ rupees.
- You get the money without paying anything from your pocket.

## What's wrong with the advert?

- The mathematics is perfect. The casino honestly computed the average of all 
profits.
- But did every player make money?
- What about the people who ended up paying at the end of the game?

```{r, echo = TRUE}
cat('Average payoff was ', round(mean(payoffs), 2), ' rupees.', sep=' ')
```

- It was a far modest amount. If the average itself is close to zero and some
people made profit then there were many who must have made loss.

## Version 2 of the game
- Start with a rupee and settle after every toss.

```{r, echo = TRUE}
starting_amount <- 1
gamble_v2 <- function(p) {
  min(cumsum(sample(
    x = c(-1, 1),
    size = n_trials,
    replace = TRUE
  ))) < starting_amount
}

bankrupt <-
  unlist(lapply(seq(n_people), gamble_v2), recursive = FALSE)
cat(length(bankrupt[bankrupt]),
    ' people will go bankrupt before the game ends.',
    sep = ' ')
```
## Survivor bias

- This sleight of hand is called survivor bias and is a result of flawed
sampling, intentionally or maliciously.
- We may have considered a toy example. But there are many serious ones.
  - Abraham Wald's analysis of aircraft surviving enemy attack in WW-II.
  - Returns advertised by mutual funds companies over a long horizon.
  - Many sports superstars being school dropouts.
  