---
title: "Analysis of chicken weight"
author: "Amey Joshi"
date: "20/11/2019"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The data describing the effect of certain choices of diet on weight of chicken
is part of every R installation. We first find out how many diet plans were
examined.
```{r}
unique(ChickWeight$Diet)
```

For each diet plan, we will see the variation in the chicken's weight from its
birth to the twentieth day. Since the steps to build the plots are very similar 
for the different diet plans, we write a function for that purpose.
```{r}
build_coplot <- function(n) {
  coplot(
  weight ~ Time |
    Chick,
  data = ChickWeight,
  given.values = as.numeric(unique(ChickWeight[ChickWeight$Diet == n,]$Chick)),
  show.given = FALSE,
  xlab = c("Time", paste("Diet", n)),
  ylab = c("Weight")
)
}
```
This is for diet 1.
```{r}
build_coplot(1)
```

Some chickens showed a modest rise in weight but many did not. A few of them 
do not have data for all twenty days, perhaps they did not survive.

For diet 2 it is
```{r}
build_coplot(2)
```

Barring one, all chickens gained weight in twenty days - some substantially, 
others more modestly.

For diet 3 it is
```{r}
build_coplot(3)
```

Every chicken gained weight, though not to the same extent.

Finally, for diet 4 the data is
```{r}
build_coplot(4)
```

The weight gain through diet 4 is monotonous in all but one case. One of the 
chickens does not have data beyong 18 days suggesting a case of mortality.

We will now analyze the efficacy of the four diets using ANOVA. In order to 
do that, we first find the weight gain for each bird.
```{r}
gain <- function(x) {
  x[length(x)] - x[1]
}

X <- with(ChickWeight, tapply(weight, Chick, gain))
weight.gain <- data.frame(Chick = names(X), gain = X)
rm(X)
```

We then merge this data with the main data set to get the weight gain by a 
bird for a diet.
```{r}
X <- merge(weight.gain, ChickWeight, by.x = "Chick", by.y = "Chick")
anova.data <- unique(X[, c("gain", "Chick", "Diet")])
rm(X)
```

Before a statistical analysis, we view the data.
```{r}
boxplot(gain ~ Diet, data = anova.data, main = "Weight gain for diet plans")
```

The means are not all the same. However, we will do an ANOVA to find out if the 
null hypothesis that all diets led to an equal weight gain is true.
```{r}
oneway.test(gain ~ Diet, data = anova.data, var.equal = TRUE)
```

The very small value of making a type I error indicates that we can reject the
null hypothesis. 

We notice that the mean gain in weight for diets 2 and 4 are quite similar. In
order to examine if there is a significant statistical difference between the two,
we will run an ANOVA on just those two data.
```{r}
anova.subset <- anova.data[anova.data$Diet %in% c(2, 4), ]
oneway.test(gain ~ Diet, data = anova.subset, var.equal = TRUE)
```

The high p-value suggests that we should not reject the null hypothesis that the
two means are indeed equal. We confirm out findings using a two-sample $t$-test.
```{r}
t.test(gain ~ Diet, data = anova.subset)
```
The p-value of the $t$-test is in line with that of ANOVA.
